---
title: "Filter DArT data with dartR"
---

This markdown provides a series of commands that may be useful when exploring and filtering DArT datasets  
It relies on the main package `dartRverse` (currently v. 1.0.6) including the packages  
`dartR.base`, `dartR.data`, and `dartR.spatial`  

Note: if you are using Ubuntu v. 24.04, you may need to separately install programs and libraries
```bash
sudo apt update && sudo apt install -y \
	cmake \
	libcurl4-openssl-dev \
	libfftw3-dev \
	libfontconfig1-dev \
	libfreetype6-dev \
	libfribidi-dev \
	libgdal-dev \
	libharfbuzz-dev \
	libjpeg-dev \
	libpng-dev \
	libssl-dev \
	libtiff5-dev \
	libudunits2-dev \
	libwebp-dev \
	libxml2-dev
```

Install necessary packages (this is for R v. 4.3)
```{r}
install.packages(c("ape", "cli", "devtools", "directlabels", "ggplot2", "leaflet",
	"leaflet.minicharts", "plotly", "remotes"))
remotes::install_version("rms", version = "6.7-1")		# or install.packages("rms") for R >= 4.4
remotes::install_version("haplo.stats", version = "1.9.7")		# or install.packages("haplo.stats") for R >= 4.4
install.packages("BiocManager")
BiocManager::install(c("SNPRelate", "snpStats"))
install.packages("dartRverse")
install.packages(c("dartR.base", "dartR.data", "dartR.spatial"))
```

Load the library and designate input files (assumed to be in the working directory, which can be set with `setwd(...)`)  
Input files are:  
1) A CSV file from DArT containing SNPs scored as "1" or "0" on two lines (alleles): `..._SNP_1.csv`  
2) A CSV file of individual metadata (header: `id,pop,lat,lon`)  
```{r}
library("dartRverse")
snps_file <- "{filename}_SNP_1.csv"
meta_file <- "meta.csv"
```

Read in the DArT data (can take a bit for larger datasets)
```{r}
# mono.rm: remove monomorphic loci
# recalc: recalculate stats in case individuals were not included in the meta_file (dropped)
# probar: progress bar
mygenl <- gl.read.dart(snps_file, ind.metafile = meta_file,
	mono.rm = TRUE, recalc = TRUE, probar = TRUE)
```

Save a copy in case of shutdown/error or for convenience later
```{r}
gl.save(mygenl, file = "input.rdata")
# to start here later, use mygenl <- gl.load("input.rdata")
```

Drop undesirable samples (from text file(s) with sample IDs, one per line)  
For example, you may want to drop low quality samples, biological replicates, and/or clones  
Clones may have been detected using DArT's distance matrix (provided with the dataset)
```{r}
# adjust these to match your files
drop_probs <- read.table("dart_drop_problematic.txt", header = FALSE)[, 1]
drop_clones <- read.table("dart_drop_clones.txt", header = FALSE)[, 1]
drop_reps <- read.table("dart_drop_replicates.txt", header = FALSE)[, 1]
drops <- c(drop_probs, drop_clones, drop_reps)

# drop
mygenl <- gl.drop.ind(mygenl, drops, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(mygenl), " individuals and ",
	nLoc(mygenl), " SNPs\n"))
snps_remain <- nLoc(mygenl)
```

# Unfiltered data
Interactively map population locations (no output)
```{r}
gl.map.interactive(mygenl)
```

Create a PDF to record various stats about the unfiltered data
```{r}
pdf("unfiltered_stats.pdf", width = 8, height = 12)
# read depth
gl.report.rdepth(mygenl)
# tag length distribution
gl.report.taglength(mygenl)
# call rate (missing data by locus or individual)
gl.report.callrate(mygenl, method = "loc")
gl.report.callrate(mygenl, method = "ind")
# reproducibility (repeatability of calls)
gl.report.reproducibility(mygenl)
# secondaries = multiple SNPs per tag
gl.report.secondaries(mygenl)
invisible(dev.off())
```

Optionally generate larger plots
```{r}
pdf("unfiltered_graphs.pdf", width = 21, height = 21)
# smear plot (can take a while)
smearplotgl <- gl.smearplot(mygenl, ind.labels = FALSE)
# minor allele frequency plot
gl.report.maf(mygenl)
invisible(dev.off())
rm(smearplotgl)
```

# Filter data to remove less reliable SNPs
Remove SNPs with low/high read depth (lower confidence in a call, higher error potentially)
```{r}
# adjust to what makes sense in the unfiltered plots above
filtered_genl <- gl.filter.rdepth(mygenl, lower = 6, upper = 100)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)

# clean up the environment to reduce memory usage
rm(mygenl)
gc()
```

Filter based on tag length; remove short loci depending on distribution
```{r}
filtered_genl <- gl.filter.taglength(filtered_genl, lower = 65, upper = 69)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Filter based on reproducibility (a proxy for problems)
```{r}
filtered_genl <- gl.filter.reproducibility(filtered_genl, threshold = 0.95)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Save a copy of the object for population diversity analyses
```{r}
gl.save(filtered_genl, file = "pop_filt_genl.rdata")
```

Drop secondary SNPs
```{r}
# can also use method = "random"
filtered_genl <- gl.filter.secondaries(filtered_genl, method = "best")
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Save a copy of the object for genetic distance estimation and evolutionary inference
```{r}
gl.save(filtered_genl, file = "dist_filt_genl.rdata")
```

# Filter for genetic structure analyses (not population diversity)
Drop outgroups if desired (from a text file with sample IDs, one per line)
```{r}
outgroups <- read.table("dart_outgroup.txt", header = FALSE)[, 1]
filtered_genl <- gl.drop.ind(filtered_genl, outgroups, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Remove SNPs found in less than a proportion of samples
```{r}
filtered_genl <- gl.filter.callrate(filtered_genl, method = "loc", threshold = 0.95)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Filter based on minor allele frequency  
(if a value > 1 is used, it is interpreted as minor allele count;  
this is preferable for a dataset of multiple populations)
```{r}
filtered_genl <- gl.filter.maf(filtered_genl, threshold = 3)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Filter out individuals with large amounts of missing data (if present)
```{r}
filtered_genl <- gl.filter.callrate(filtered_genl, method = "ind", threshold = 0.50)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Remove fully NA and monomorphic loci (if any are left following sample removal)
```{r}
filtered_genl <- gl.filter.allna(filtered_genl)
filtered_genl <- gl.filter.monomorphs(filtered_genl)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

To further avoid linkage, you can use a Hamming filter to remove tags that are too similar  
and may be the same locus. The threshold (default 0.2) may need to be adjusted  
compared to the length of a tag (69 bp). Assume that less than c. 15% difference  
or fewer than c. 10 bp differences between tags means they represent the same locus.  
This can take a long time (not recommended if more than 20 k SNPs remain)
```{r}
filtered_genl <- gl.filter.hamming(filtered_genl, tag.length = 69, rs = 5,
	threshold = 0.15, pb = TRUE)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Save the object for import later with `gl.load`
```{r}
gl.save(filtered_genl, file = "filtered_genl.rdata")

# clean up memory
rm(filtered_genl)
gc()
```

# ANALYSES

## Genetic structure
Load the dataset
```{r}
filtered_genl <- gl.load(file = "filtered_genl.rdata")
snps_remain <- nLoc(filtered_genl)
```

### PCA
Drop outgroups if desired or not done already (from a text file with sample IDs, one per line)
```{r}
outgroups <- read.table("dart_outgroup.txt", header = FALSE)[, 1]
filtered_genl <- gl.drop.ind(filtered_genl, outgroups, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Run a PCA of the genlight SNPs
```{r}
pca <- gl.pcoa(filtered_genl, nfactors = 5, parallel = TRUE, n.cores = 8)
```

Plot with dartR
```{r}
pdf("pca.pdf")
cat(paste0("Plotting a PCA based on ", snps_remain, " SNPs\n"))
plot1 <- gl.pcoa.plot(pca, filtered_genl, pop.labels = "pop", label.size = 0.8,
	xaxis = 1, yaxis = 2, zaxis = NULL, interactive = FALSE)
plot2 <- gl.pcoa.plot(pca, filtered_genl, pop.labels = "pop", label.size = 0.8,
	xaxis = 1, yaxis = 3, zaxis = NULL, interactive = FALSE)
plot3 <- gl.pcoa.plot(pca, filtered_genl, pop.labels = "pop", label.size = 0.8,
	xaxis = 1, yaxis = 4, zaxis = NULL, interactive = FALSE)
invisible(dev.off())
```

Or plot interactively
```{r}
plot4 <- gl.pcoa.plot(pca, filtered_genl, pop.labels = "pop",
	xaxis = 1, yaxis = 2, zaxis = NULL, interactive = TRUE)
```
Or plot interactively in 3D
```{r}
plot5 <- gl.pcoa.plot(pca, filtered_genl, pop.labels = "pop",
	xaxis = 1, yaxis = 2, zaxis = 3, interactive = TRUE)
```

### Structure
Drop outgroups if desired or not done already (from a text file with sample IDs, one per line)
```{r}
outgroups <- read.table("dart_outgroup.txt", header = FALSE)[, 1]
filtered_genl <- gl.drop.ind(filtered_genl, outgroups, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Output a file formatted for Structure, but first replace pop with a number in the genlight
```{r}
filtered_genl <- gl.sort(filtered_genl, sort.by = "ind")
index <- 1
pops <- sort(unique(filtered_genl$pop))
for (pop in pops) {
	levels(filtered_genl$pop)[match(pop, levels(filtered_genl$pop))] <- index
	index <- index + 1
}
filtered_genl <- gl.sort(filtered_genl, sort.by = "pop")

gl2structure(filtered_genl, outfile = "structure_input.str", add.columns = filtered_genl$pop,
	export.marker.names = FALSE, outpath = ".")

coding <- cbind(filtered_genl$ind.names, filtered_genl$pop)
write.table(coding, file = "str_pops.txt", row.names = FALSE,
	col.names = FALSE, quote = FALSE, sep = "\t")
```

Clean up memory
```{r}
rm(filtered_genl)
gc()
```

## Distance
Load the dataset
```{r}
dist_filt_genl <- gl.load(file = "dist_filt_genl.rdata")
```

Drop outgroups if desired (from a text file with sample IDs, one per line)
```{r}
outgroups <- read.table("dart_outgroup.txt", header = FALSE)[, 1]
dist_filt_genl <- gl.drop.ind(dist_filt_genl, outgroups, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(dist_filt_genl), " individuals and ",
	nLoc(dist_filt_genl), " SNPs\n"))
```

Output files for analysis and display with SplitsTree
```{r}
# method 3 = concatenate SNPs with ambiguities used for heterozygous sites
gl2fasta(dist_filt_genl, outfile = "dist_filtered_SNPs.fasta", method = 3, outpath = ".")
```

Install and load required libraries to generate distance matrices from DNA sequences
```{r}
library("devtools")
install_github("simjoly/pofadinr")
```

Generate a distance matrix and write to file
```{r}
library("ape")
library("pofadinr")
mydnabin <- fasta2DNAbin("dist_filtered_SNPs.fasta")
temp <- as.character(mydnabin)
temp[temp == "n"] <- "?"
mydnabin <- as.DNAbin(temp)
distance <- dist.snp(mydnabin, model = "MATCHSTATES", as.matrix = TRUE)
taxa <- rownames(distance)
taxa_block <- paste0("BEGIN TAXA;\n\tDIMENSIONS NTAX=", length(taxa), ";\n\t",
	"TAXLABELS ", paste(taxa, collapse = " "), ";\nEND;\n")
dist_block <- paste0("BEGIN DISTANCES;\n\tFORMAT\n\t\tTRIANGLE=BOTH\n\t\tDIAGONAL\n\t\t",
	"LABELS=LEFT\n\t;\n\tMATRIX\n")
outfile <- file("distances_MATCHSTATES.nex", open = "w")
writeLines("#NEXUS", con = outfile)
writeLines(taxa_block, con = outfile)
writeLines(dist_block, con = outfile)
write.table(as.matrix(distance), file = outfile, col.names = FALSE,
	append = TRUE, quote = FALSE)
writeLines("\t;\nEND;\n", con = outfile)
close(outfile)
```

Clean up memory
```{r}
rm(temp, mydnabin, dist_filt_genl)
gc()
```

## Evolutionary inference

### SPEEDEMON
Load the distance dataset (with outgroups; no secondaries)
```{r}
snapper_genl <- gl.load("dist_filt_genl.rdata")
```

Filter the genlight based on minor allele count, then downsample if necessary  
To keep specific samples, put IDs in a text file `dart_speedemon_samples.txt`, one per line
```{r}
snapper_genl <- gl.filter.maf(snapper_genl, threshold = 3)
keep <- read.table("dart_speedemon_samples.txt", header = FALSE)[, 1]
snapper_genl <- gl.keep.ind(snapper_genl, as.character(keep), mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(snapper_genl), " individuals and ",
	nLoc(snapper_genl), " SNPs\n"))
```

Filter for high coverage and generate output removing SNPs only variable in a single population
```{r}
snapper_genl <- gl.filter.callrate(snapper_genl, method = "loc", threshold = 1)
gl2snapper(snapper_genl, outfile = "dart_speedemon_snapp.nex", outpath = ".",
	rm.autapomorphies = TRUE, nloc = 1000)
```
Note: BEAUti may have issues reading this in, so you may have to manually change `integerdata` to `snp` in the output file  

Create a similar output of binary characters to check that relationships make sense in SplitsTree
```{r}
# from the gl2snapper function
freqs <- gl.allele.freq(snapper_genl, by = "popxloc")
subset <- freqs[freqs$frequency < 0.99, ]
subset <- subset[subset$frequency > 1e-04, ]
tbl <- table(subset$locus)
single_count_loci <- names(tbl[tbl == 1])
mygenl <- gl.drop.loc(snapper_genl, loc.list = single_count_loci)

# subsample and change labels for output
mygenl <- gl.subsample.loci(mygenl, 1000, method = "random")
mymat <- as.matrix(mygenl)
labels <- vector(length = length(rownames(mymat)))
for (index in seq_len(length(rownames(mymat)))) {
	labels[index] <- paste0(pop(mygenl)[index], "_", indNames(mygenl)[index])
}
rownames(mymat) <- labels

# generate output
mymat[is.na(mymat)] <- "?"
mymat[mymat == "0"] <- "00"
mymat[mymat == "1"] <- "01"
mymat[mymat == "2"] <- "11"
matrix_block <- apply(mymat, 1, function(x) paste0(x, collapse = ""))
data_block <- paste0("BEGIN TAXA;\n\tDIMENSIONS NTAX=", nrow(mymat), ";\n\t",
	"TAXLABELS ", paste0(rownames(mymat), collapse = " "), ";\nEND;\n",
	"BEGIN CHARACTERS;\n\tDIMENSIONS NCHAR=", 2 * ncol(mymat), ";\n\t",
	"FORMAT\n\t\tDATATYPE=STANDARD\n\t\tMISSING=?\n\t\t",
	"LABELS=NO\n\t\tSYMBOLS=\"01\";\n\tMATRIX")
outfile <- file("dart_speedemon_splits.nex", open = "w")
writeLines("#NEXUS", con = outfile)
writeLines(data_block, con = outfile)
for (entry in matrix_block) {
	writeLines(paste0(c("\t\t", entry), collapse = ""), con = outfile)
}
writeLines("\t;\nEND;", con = outfile)
close(outfile)
```

Clean up memory
```{r}
rm(snapper_genl)
gc()
```

## Population genetics

### Pairwise Fst and IBD
Load the dataset without linkage
```{r}
filtered_genl <- gl.load(file = "filtered_genl.rdata")
```

Check sample coverage
```{r}
summary(filtered_genl$pop)
```

Drop pops if needed (e.g. too few samples)  
Can list them manually (e.g. `drops <- c("pop1", "pop2", ...)`),   
or use a text file `dart_drop_pops.txt`, one pop per line
```{r}
# adjust as necessary
drops <- read.table("dart_drop_pops.txt", header = FALSE)[, 1]

filtered_genl <- gl.drop.pop(filtered_genl, drops, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
```

Drop specific individuals for even population sampling (e.g. five per pop)  
If wanting specific samples, put the sample IDs of samples to drop in a text file `dart_drop_samples.txt`
```{r}
drops <- read.table("dart_drop_samples.txt", header = FALSE)[, 1]
filtered_genl <- gl.drop.ind(filtered_genl, drops, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
```
**OR** if wanting random drops  
```{r}
filtered_genl <- gl.subsample.ind(filtered_genl, 5, by.pop = TRUE, replace = FALSE)
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
```

Calculate pairwise Fst with StAMPP and write to file
```{r}
fsts <- gl.fst.pop(filtered_genl, nboots = 1)
write.table(fsts, file = "StAMPP_Fst.txt", quote = FALSE, row.names = TRUE)
```

Run isolation by distance and create a pdf if interested (uses STaMPP Fst by default)
```{r}
pdf("StAMPP_IBD_all.pdf", width = 7, height = 7)
gl.ibd(filtered_genl)
invisible(dev.off())
```

Alternatively, subset to the populations of interest for IBD  
(Repeat these steps as necessary)  
Put the pops to drop in a `drop_pops.txt` file
```{r}
drops <- read.table("drop_pops.txt", header = FALSE)[, 1]
```
**OR** select from the list of pops
```{r}
library(stringr)
pops_avail <- as.character(unique(filtered_genl$pop))
# change as necessary
target <- "pop_string"

keep_indices <- grep(target, pops_avail)
drops <- pops_avail[-keep_indices]
```

Subset
```{r}
genl_subset <- gl.drop.pop(filtered_genl, drops, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(genl_subset), " individuals and ",
	nLoc(genl_subset), " SNPs\n"))
```

Run isolation by distance for the subset
```{r}
pdf("StAMPP_IBD_subset.pdf", width = 7, height = 7)
gl.ibd(genl_subset)
invisible(dev.off())
```

Clean up memory
```{r}
rm(filtered_genl, genl_subset)
gc()
```

### Diversity
Load the dataset
```{r}
pop_filt_genl <- gl.load(file = "pop_filt_genl.rdata")
```

Drop undesirable samples such as outgroups or hybrids if present  
Put samples IDs to drop in appropriate text files, one per line
```{r}
# adjust as necessary
outgroups <- read.table("dart_outgroup.txt", header = FALSE)[, 1]
hybrids <- read.table("dart_hybrids.txt", header = FALSE)[, 1]
drops <- c(outgroups, hybrids)

pop_filt_genl <- gl.drop.ind(pop_filt_genl, drops, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(pop_filt_genl), " individuals and ",
	nLoc(pop_filt_genl), " SNPs\n"))
```

Check sample coverage
```{r}
summary(pop_filt_genl$pop)
```

Drop pops if needed (e.g. too few samples)  
Can list them manually (e.g. `drops <- c("pop1", "pop2", ...)`),   
or use a text file `dart_drop_pops.txt`, one pop per line
```{r}
# adjust as necessary
drops <- read.table("dart_drop_pops.txt", header = FALSE)[, 1]

pop_filt_genl <- gl.drop.pop(pop_filt_genl, drops, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(pop_filt_genl), " individuals and ",
	nLoc(pop_filt_genl), " SNPs\n"))
```

Drop individuals for even population sampling (e.g. five per pop)  
If wanting specific samples, put the sample IDs of samples to drop in a text file `dart_drop_samples.txt`
```{r}
drops <- read.table("dart_drop_samples.txt", header = FALSE)[, 1]
pop_filt_genl <- gl.drop.ind(pop_filt_genl, drops, mono.rm = FALSE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(pop_filt_genl), " individuals and ",
	nLoc(pop_filt_genl), " SNPs\n"))
```
**OR** if wanting random drops  
```{r}
pop_filt_genl <- gl.subsample.ind(pop_filt_genl, 5, by.pop = TRUE, replace = FALSE)
cat(paste0("Filtered to a genlight of ", nInd(pop_filt_genl), " individuals and ",
	nLoc(pop_filt_genl), " SNPs\n"))
```

If your dataset comprises multiple taxa, consider also subsetting by taxon  
It is not clear if this is the best way to handle multi-species datasets, or if  
those datasets should be otherwise included in diversity metric estimates  
Specify the taxon of each pop using a text file `dart_pop_taxa.txt` with two columns,  
population and taxon, tab separated and one per line with no header
```{r}
pop_taxa <- read.table("dart_pop_taxa.txt", sep = "\t", header = FALSE)
```

#### Autosomal/SNP heterozygosity
To ensure no missing data, each population needs to be filtered separately  
Only loci found in all samples in that population will be kept (zero missing data)  
Note: currently, dartR does not report Fis with autosomal adjustment(!) (i.e. including secondaries)
```{r}
stde <- function(x) {
	sqrt(var(x, na.rm = TRUE) / sum(!is.na(x)))
}

pops <- sort(unique(pop_filt_genl$pop))
summary <- as.data.frame(matrix(ncol = 14, nrow = length(pops)))
colnames(summary) <- c("Samples", "Loci_auto", "SNPs", "Ho_auto", "Ho_auto_se", "He_auto", "He_auto_se",
	"unlink_SNPs", "Ho_SNP", "Ho_SNP_se", "uHe_SNP", "uHe_SNP_se", "Fis", "Fis_se")
rownames(summary) <- pops

# subset by taxon, if that information is present
if (exists("pop_taxa")) {
	taxa <- unique(pop_taxa[, 2])
	for (taxon in taxa) {
		keep_pops <- pop_taxa[, 1][pop_taxa[, 2] == taxon]
		keep_pops <- keep_pops[keep_pops %in% pops]
		if (length(keep_pops) > 0) {
			input_genl <- gl.keep.pop(pop_filt_genl, keep_pops, mono.rm = TRUE, recalc = TRUE)
			get_invariant <- gl.report.secondaries(input_genl)
			taxon_loci <- get_invariant[1, 2]
			taxon_invariant <- get_invariant[5, 2]		# observed invariant rather than estimated
			# cycle through the pops of that taxon
			for (pop in keep_pops) {
				# subset and estimate autosomal stats
				sub_genl <- gl.keep.pop(input_genl, pop, mono.rm = FALSE, recalc = FALSE)
				sub_genl <- gl.filter.allna(sub_genl, recalc = TRUE)		# remove all NA loci
				sub_genl <- gl.filter.callrate(sub_genl, method = "loc", threshold = 1)		# remove partly missing
				get_invariant <- gl.report.secondaries(sub_genl)
				pop_loci <- get_invariant[1, 2]
				stats <- gl.report.heterozygosity(sub_genl, n.invariant = (pop_loci / taxon_loci) * taxon_invariant)
				summary[pop, "Samples"] <- nInd(sub_genl)
				summary[pop, "Loci_auto"] <- pop_loci
				summary[pop, "SNPs"] <- stats[1, "n.Loc"]
				summary[pop, "Ho_auto"] <- stats[1, "Ho.adj"]
				summary[pop, "Ho_auto_se"] <- stats[1, "Ho.adjSE"]
				summary[pop, "He_auto"] <- stats[1, "He.adj"]
				summary[pop, "He_auto_se"] <- stats[1, "He.adjSE"]

				# estimate stats from unlinked variable SNPs
				sub_genl <- gl.filter.monomorphs(sub_genl)
				sub_genl <- gl.filter.secondaries(sub_genl, method = "random")
				snp_stats <- gl.report.heterozygosity(sub_genl, n.invariant = 0)
				summary[pop, "unlink_SNPs"] <- snp_stats[1, "n.Loc"]
				summary[pop, "Ho_SNP"] <- snp_stats[1, "Ho"]
				summary[pop, "Ho_SNP_se"] <- snp_stats[1, "HoSE"]
				summary[pop, "uHe_SNP"] <- snp_stats[1, "uHe"]
				summary[pop, "uHe_SNP_se"] <- snp_stats[1, "uHeSE"]
				summary[pop, "Fis"] <- snp_stats[1, "FIS"]
				summary[pop, "Fis_se"] <- snp_stats[1, "FISSE"]
			}
		}
	}
} else {		# otherwise, run for each pop separately
	input_genl <- pop_filt_genl
	get_invariant <- gl.report.secondaries(input_genl)
	dataset_loci <- get_invariant[1, 2]
	dataset_invariant <- get_invariant[5, 2]		# observed invariant rather than estimated
	for (pop in pops) {
		# subset and estimate autosomal stats
		sub_genl <- gl.keep.pop(input_genl, pop, mono.rm = FALSE, recalc = FALSE)
		sub_genl <- gl.filter.allna(sub_genl, recalc = TRUE)		# remove all NA loci
		sub_genl <- gl.filter.callrate(sub_genl, method = "loc", threshold = 1)		# remove partly missing
		get_invariant <- gl.report.secondaries(sub_genl)
		pop_loci <- get_invariant[1, 2]
		stats <- gl.report.heterozygosity(sub_genl, n.invariant = (pop_loci / dataset_loci) * dataset_invariant)
		summary[pop, "Samples"] <- nInd(sub_genl)
		summary[pop, "Loci_auto"] <- get_invariant[1, 2]
		summary[pop, "SNPs"] <- stats[1, "n.Loc"]
		summary[pop, "Ho_auto"] <- stats[1, "Ho.adj"]
		summary[pop, "Ho_auto_se"] <- stats[1, "Ho.adjSE"]
		summary[pop, "He_auto"] <- stats[1, "He.adj"]
		summary[pop, "He_auto_se"] <- stats[1, "He.adjSE"]

		# estimate stats from unlinked variable SNPs
		sub_genl <- gl.filter.monomorphs(sub_genl)
		sub_genl <- gl.filter.secondaries(sub_genl, method = "random")
		snp_stats <- gl.report.heterozygosity(sub_genl, n.invariant = 0)
		summary[pop, "unlink_SNPs"] <- snp_stats[1, "n.Loc"]
		summary[pop, "Ho_SNP"] <- snp_stats[1, "Ho"]
		summary[pop, "Ho_SNP_se"] <- snp_stats[1, "HoSE"]
		summary[pop, "uHe_SNP"] <- snp_stats[1, "uHe"]
		summary[pop, "uHe_SNP_se"] <- snp_stats[1, "uHeSE"]
		summary[pop, "Fis"] <- snp_stats[1, "FIS"]
		summary[pop, "Fis_se"] <- snp_stats[1, "FISSE"]
	}
}
```

Write to file
```{r}
write.table(summary, file = paste0("diversity_stats.txt"),
	quote = FALSE, row.names = TRUE)
```

Plot diversity stats to pdf
```{r}
mybarplot <- function(data_val, data_se, names, main) {
	myplot <- barplot(data_val, names = names, las = 2,
		main = main,
		ylim = c(min(c(0, 1.2 * (data_val - data_se * 2))),
			max(0, 1.2 * max(data_val + data_se * 2))))
	arrows(myplot, y0 = data_val + data_se * 2, y1 = data_val - data_se * 2,
		angle = 90, code = 3, length = 0.1)
}

pdf("diversity_stats.pdf", width = 10, height = 10)
mybarplot(summary$Ho_auto, summary$Ho_auto_se, names = rownames(summary),
	main = "Autosomal observed heterozygosity (Ho)")
mybarplot(summary$Ho_SNP, summary$Ho_SNP_se, names = rownames(summary),
	main = "SNP observed heterozygosity (Ho)")
mybarplot(summary$He_auto, summary$He_auto_se, names = rownames(summary),
	main = "Autosomal expected heterozygosity (He)")
mybarplot(summary$uHe_SNP, summary$uHe_SNP_se, names = rownames(summary),
	main = "SNP unbiased expected heterozygosity (uHe)")
mybarplot(summary$Fis, summary$Fis_se, names = rownames(summary),
	main = "Inbreeding coefficient Fis\n(1 - Ho / uHe)")
invisible(dev.off())
```
