---
title: "Filter DArT data with dartR"
---

This markdown provides a series of commands that may be useful when exploring and filtering DArT datasets  
It relies on the package `dartRverse` (using v. 1.0.6) including the packages  
`dartR.base` (v. 1.0.8), `dartR.data` (v. 1.0.2), and `dartR.spatial` (v. 0.78)  

Note: if you are using Ubuntu v. 24.04, you may need to separately install libraries:
```bash
sudo apt update && apt install -y \
	libcurl4-openssl-dev \
	libfftw3-dev \
	libfontconfig1-dev \
	libfreetype6-dev \
	libfribidi-dev \
	libgdal-dev \
	libharfbuzz-dev \
	libjpeg-dev \
	libpng-dev \
	libssl-dev \
	libtiff5-dev \
	libwebp-dev \
	libxml2-dev
```

Install necessary packages (this is for R v. 4.3)
```{r}
install.packages(c("ape", "devtools", "directlabels", "plotly", "remotes"))
remotes::install_version("rms", version = "6.7-1")
remotes::install_version("haplo.stats", version = "1.9.7")
install.packages("BiocManager")
BiocManager::install(c("SNPRelate", "snpStats"))
install.packages("dartRverse")
install.packages(c("dartR.base", "dartR.data", "dartR.spatial"))
```

Load the library and designate input files (assumed to be in the working directory, which can be set with `setwd(...)`  
Input files are:  
1) A CSV file from DArT containing SNPs scored as "1" or "0" on two lines (alleles): `..._SNP_1.csv`  
2) A CSV file of individual metadata (header: `id,pop,lat,lon`)  
```{r}
suppressMessages(library("dartRverse"))
snps_file <- "*_SNP_1.csv"
meta_file <- "meta.csv"
```

Read in the DArT data (can take a bit for larger datasets)
```{r}
# mono.rm: remove monomorphic loci
# recalc: recalculate stats in case individuals were not included in the meta_file (dropped)
# probar: progress bar
mygenl <- gl.read.dart(snps_file, ind.metafile = meta_file,
	mono.rm = TRUE, recalc = TRUE, probar = TRUE)
```

Save a copy in case of shutdown/error or for convenience later
```{r}
gl.save(mygenl, file = "input.rdata")
```

If wanting to start again from this point later, start here
```{r}
mygenl <- gl.load("input.rdata")
```

Drop outgroups if desired (from a text file with sample IDs, one per line)
```{r}
outgroups <- read.table("outgroup_samples.txt", header = FALSE)[, 1]
mygenl <- gl.drop.ind(mygenl, outgroups, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(mygenl), " individuals and ",
	nLoc(mygenl), " SNPs\n"))
snps_remain <- nLoc(mygenl)
```

# Unfiltered data
Interactively map population locations (no output)
```{r}
gl.map.interactive(mygenl)
```

Create a PDF to record various stats about the unfiltered data
```{r}
pdf("unfiltered_stats.pdf", width = 8, height = 12)
# read depth
gl.report.rdepth(mygenl)
# tag length distribution
gl.report.taglength(mygenl)
# call rate (missing data by locus or individual)
gl.report.callrate(mygenl, method = "loc")
gl.report.callrate(mygenl, method = "ind")
# reproducibility (repeatability of calls)
gl.report.reproducibility(mygenl)
# secondaries = multiple SNPs per tag
gl.report.secondaries(mygenl)
invisible(dev.off())
```

Optionally generate larger plots
```{r}
pdf("unfiltered_graphs.pdf", width = 21, height = 21)
# smear plot (can take a while)
smearplotgl <- gl.smearplot(mygenl, ind.labels = FALSE)
# minor allele frequency plot
gl.report.maf(mygenl)
invisible(dev.off())
```

# Filter data to remove less reliable SNPs
Remove SNPs with low/high read depth (lower confidence in a call, higher error potentially)
```{r}
filtered_genl <- gl.filter.rdepth(mygenl, lower = 6, upper = 100)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)

# clean up the environment to reduce memory usage
rm(mygenl)
gc()
```

Filter based on tag length; remove short loci depending on distribution
```{r}
filtered_genl <- gl.filter.taglength(filtered_genl, lower = 60, upper = 69)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Filter based on reproducibility (a proxy for problems)
```{r}
filtered_genl <- gl.filter.reproducibility(filtered_genl, threshold = 0.95)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Make a copy of the object for population diversity analyses (save it for later if needed)
```{r}
pop_filt_genl <- filtered_genl
gl.save(pop_filt_genl, file = "pop_filt_genl.rdata")
```

# Filter for genetic structure analyses (not population diversity)
Remove SNPs found in less than a proportion of samples
```{r}
filtered_genl <- gl.filter.callrate(filtered_genl, method = "loc", threshold = 0.95)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Filter based on minor allele frequency  
(if a value > 1 is used, it is interpreted as minor allele count;  
this is preferable for a dataset of multiple populations)
```{r}
filtered_genl <- gl.filter.maf(filtered_genl, threshold = 3)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Filter out individuals with large amounts of missing data (if present)
```{r}
filtered_genl <- gl.filter.callrate(filtered_genl, method = "ind", threshold = 0.50)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Remove monomorphic and fully NA loci (if any are left)
```{r}
filtered_genl <- gl.filter.monomorphs(filtered_genl)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Drop secondary SNPs
```{r}
# can also use metho = "random"
filtered_genl <- gl.filter.secondaries(filtered_genl, method = "best")
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

To avoid linkage, use a Hamming filter to remove tags that are too similar (the same locus)  
The threshold (default 0.2) needs to consider the length of a tag (69 bp)  
Assume that if there are fewer than c. 12% difference or c. 8 bp differences, it is the same locus  
This can take a long time (not recommended if > 20 k SNPs remain; may have to run overnight)
```{r}
filtered_genl <- gl.filter.hamming(filtered_genl, tag.length = 69, rs = 5,
	threshold = 0.12, pb = TRUE)
diff <- snps_remain - nLoc(filtered_genl)
cat(paste0("Removed ", diff, " SNPs\n"))
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Save the object for import later with `gl.load`
```{r}
gl.save(filtered_genl, file = "filtered_genl.rdata")
# free up memory
gc()
```

# ANALYSES

## Genetic structure

### PCA
Run a PCA of the genlight SNPs
```{r}
pca <- gl.pcoa(filtered_genl, nfactors = 5, parallel = TRUE, n.cores = 8)
```

Plot with dartR
```{r}
pdf("pca.pdf")
gl.pcoa.plot(pca, filtered_genl, pop.labels = "pop",
	xaxis = 1, yaxis = 2, zaxis = NULL, interactive = FALSE)
cat(paste0("Plotting a PCA based on ", snps_remain, " SNPs\n"))
invisible(dev.off())
```

Or plot interactively
```{r}
gl.pcoa.plot(pca, filtered_genl, pop.labels = "pop", xaxis = 1, yaxis = 2,
	zaxis = NULL, interactive = TRUE)
```
Or plot interactively in 3D
```{r}
gl.pcoa.plot(pca, filtered_genl, pop.labels = "pop", xaxis = 1, yaxis = 2,
	zaxis = 3, interactive = TRUE)
```

### Distance
Output files for analysis and display with SplitsTree
```{r}
# method 3 = concatenate SNPs with ambiguities used for heterozygous sites
gl2fasta(filtered_genl, outfile = "filtered_SNPs.fasta", method = 3, outpath = ".")
```

```{r}
# install and load required libraries to generate distance matrices from DNA sequences
library("devtools")
install_github("simjoly/pofadinr")
```

```{r}
library("ape")
library("pofadinr")
mydnabin <- fasta2DNAbin("filtered_SNPs.fasta")
temp <- as.character(mydnabin)
temp[temp == "n"] <- "?"
mydnabin <- as.DNAbin(temp)
distance <- dist.snp(mydnabin, model = "MATCHSTATES", as.matrix = TRUE)
outfile <- file("distances_MATCHSTATES.txt", open = "w")
writeLines(as.character(nInd(filtered_genl)), con = outfile)
write.table(distance, file = outfile, sep = " ", append = TRUE,
	col.names = FALSE, quote = FALSE)
close(outfile)
rm(temp)
```

### Structure  
Output a file formatted for Structure, but first replace pop with a number in the genlight
```{r}
strgenl <- filtered_genl
index <- 1
pops <- sort(unique(strgenl$pop))
for (pop in pops) {
	levels(strgenl$pop)[match(pop, levels(strgenl$pop))] <- index
	index <- index + 1
}
gl2structure(strgenl, outfile = "structure_SNPs.str", add.columns = strgenl$pop,
	export.marker.names = FALSE, outpath = ".")
coding <- cbind(strgenl$ind.names, strgenl$pop)
write.table(coding[order(as.numeric(coding[, 2])), ], file = "structure_pops.txt",
	row.names = FALSE, col.names = FALSE, quote = FALSE)

# clean up
rm(strgenl)
gc()
```

### SNAPPER  
Load the filtered dataset (secondaries already removed), then downsample and remove monomorphic  
Filter for high coverage, then generate the output removing SNPs only variable in a single population
```{r}
snapper_genl <- gl.load("filtered_genl.rdata")
keep <- read.table("snapper_samples.txt", header = FALSE)[, 1]
snapper_genl <- gl.keep.ind(snapper_genl, as.character(keep), mono.rm = TRUE, recalc = TRUE)
snapper_genl <- gl.filter.callrate(snapper_genl, method = "loc", threshold = 1)
gl2snapper(snapper_genl, outfile = "snapper_snapp.nex", outpath = ".", rm.autapomorphies = TRUE, nloc = 1000)
```
Note: BEAUti may have issues reading this in, so you may have to change `integerdata` to `snp` in the output file  

NEXUS of binary characters  
This could be useful for SplitsTree and uncorrected P-distances  
As an example, use the `snapper_genl` just filtered above (or modify for `filtered_genl`)  
If interested in changing display labels, read in a text file with two columns (original and new), no header  
```{r}
mygenl <- gl.subsample.loci(snapper_genl, 1000, method = "random")
mymat <- as.matrix(mygenl)

# rename if desired (comment out if not)
names <- read.table("rename.txt", header = FALSE, sep = "\t")
names$V1 <- as.character(names$V1)
for (index in seq_len(nrow(mymat))) {
	rownames(mymat)[index] <- names$V2[match(rownames(mymat)[index], names$V1)]
}

# generate output
mymat[is.na(mymat)] <- "?"
mymat[mymat == "0"] <- "00"
mymat[mymat == "1"] <- "01"
mymat[mymat == "2"] <- "11"
matrix_block <- apply(mymat, 1, function(x) paste0(x, collapse = ""))
data_block <- paste0("BEGIN TAXA;\n\tDIMENSIONS NTAX=", nrow(mymat), ";\n\t",
	"TAXLABELS ", paste0(rownames(mymat), collapse = " "), ";\nEND;\n",
	"BEGIN CHARACTERS;\n\tDIMENSIONS NCHAR=", 2 * ncol(mymat), ";\n\t",
	"FORMAT\n\t\tDATATYPE=STANDARD\n\t\tMISSING=?\n\t\t",
	"LABELS=NO\n\t\tSYMBOLS=\"01\";\n\tMATRIX")
outfile <- file("snapper_splits.nex", open = "w")
writeLines("#NEXUS", con = outfile)
writeLines(data_block, con = outfile)
for (entry in matrix_block) {
	writeLines(paste0(c("\t\t", entry), collapse = ""), con = outfile)
}
writeLines("\t;\nEND;", con = outfile)
close(outfile)

# clean up
rm(snapper_genl)
gc()
```

### Pairwise Fst and IBD
Load the dataset (if saved earlier and starting here)
```{r}
filtered_genl <- gl.load(file = "filtered_genl.rdata")
```

Drop individuals for even population sampling (e.g. six per pop)  
Put the sample IDs of samples to drop to leave only six per pop in a text file `drop_samples_fst.txt`
```{r}
drops <- read.table("drop_samples_fst.txt", header = FALSE)[, 1]
filtered_genl <- gl.drop.ind(filtered_genl, drops, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl), " individuals and ",
	nLoc(filtered_genl), " SNPs\n"))
snps_remain <- nLoc(filtered_genl)
```

Calculate pairwise Fst with StAMPP and write to file
```{r}
fsts <- gl.fst.pop(filtered_genl, nboots = 1)
### export the pairwise Fst values to file
write.table(fsts, file = "StAMPP_Fst.txt",
	quote = FALSE, row.names = TRUE)
```

Run IBD if interested (will use STaMPP Fst by default)
```{r}
pdf("StAMPP_IBD_all.pdf", width = 7, height = 7)
gl.ibd(filtered_genl)
invisible(dev.off())
```

Alternatively, subset to the populations of interest for IBD  
Put the pops to drop in a `drop_pops.txt` file
```{r}
drops <- read.table("drop_pops.txt", header = FALSE)[, 1]
```
OR, select from the list of pops
```{r}
library(stringr)
pops_avail <- as.character(unique(filtered_genl$pop))
target <- "pop_string"
keep_indices <- grep(target, pops_avail)
drops <- pops_avail[-keep_indices]
```

```{r}
filtered_genl_subset <- gl.drop.pop(filtered_genl, drops, mono.rm = TRUE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(filtered_genl_subset), " individuals and ",
	nLoc(filtered_genl_subset), " SNPs\n"))
```

```{r}
pdf("StAMPP_IBD_subset.pdf", width = 7, height = 7)
gl.ibd(filtered_genl_subset)
invisible(dev.off())
```

# Popgen diversity
Load the dataset (if saved earlier and starting here)
```{r}
pop_filt_genl <- gl.load(file = "pop_filt_genl.rdata")
```

Check sample coverage
```{r}
summary(pop_filt_genl$pop)
```

Drop individuals so that populations have even coverage for comparison (e.g. five samples per pop)  
Put the sample IDs of samples to drop to leave only five per pop in a text file `drop_samples_diversity.txt`
```{r}
drops <- read.table("drop_samples_diversity.txt", header = FALSE)[, 1]
pop_filt_genl <- gl.drop.ind(pop_filt_genl, drops, mono.rm = FALSE, recalc = TRUE)
cat(paste0("Filtered to a genlight of ", nInd(pop_filt_genl), " individuals and ",
	nLoc(pop_filt_genl), " SNPs\n"))
snps_remain <- nLoc(pop_filt_genl)
```

Check sample coverage
```{r}
summary(pop_filt_genl$pop)
```

## Autosomal/SNP heterozygosity  
To ensure no missing data, each population needs to be filtered separately  
Only loci found in all samples in that population will be kept (zero missing data)  
Note: currently, the function will not report Fis for autosomal(!)
```{r}
stde <- function(x) {
	sqrt(var(x, na.rm = TRUE) / sum(!is.na(x)))
}

pops <- sort(unique(pop_filt_genl$pop))
summary <- as.data.frame(matrix(ncol = 15, nrow = length(pops)))
colnames(summary) <- c("Samples", "Loci_auto", "Ho_auto", "Ho_auto_se", "He_auto", "He_auto_se",
	"Loci_SNPs", "Poly_SNPs", "Mono_SNPs", "Ho_SNP", "Ho_SNP_se", "He_SNP", "He_SNP_se", "Fis", "Fis_se")
rownames(summary) <- pops
for (pop in pops) {
	# subset and calculate autosomal stats
	sub_genl <- gl.keep.pop(pop_filt_genl, pop, mono.rm = FALSE, recalc = TRUE)
	sub_genl <- gl.filter.callrate(sub_genl, method = "loc", threshold = 1)
	get_invariant <- gl.report.secondaries(sub_genl)
	stats <- gl.report.heterozygosity(sub_genl, n.invariant = get_invariant[7, 2])
	summary[pop, "Samples"] <- nInd(sub_genl)
	summary[pop, "Loci_auto"] <- stats[1, "n.Loc"]
	summary[pop, "Ho_auto"] <- stats[1, "Ho.adj"]
	summary[pop, "Ho_auto_se"] <- stats[1, "Ho.adjSE"]
	summary[pop, "He_auto"] <- stats[1, "He.adj"]
	summary[pop, "He_auto_se"] <- stats[1, "He.adjSE"]

	# calculate stats from the equivalent unlinked SNPs
	filt_sub_genl <- gl.filter.secondaries(sub_genl, method = "best")
	snp_stats <- gl.report.heterozygosity(filt_sub_genl, n.invariant = 0)
	summary[pop, "Loci_SNPs"] <- snp_stats[1, "n.Loc"]
	summary[pop, "Poly_SNPs"] <- snp_stats[1, "polyLoc"]
	summary[pop, "Mono_SNPs"] <- snp_stats[1, "monoLoc"]
	summary[pop, "Ho_SNP"] <- snp_stats[1, "Ho"]
	summary[pop, "Ho_SNP_se"] <- snp_stats[1, "HoSE"]
	summary[pop, "He_SNP"] <- snp_stats[1, "uHe"]
	summary[pop, "He_SNP_se"] <- snp_stats[1, "uHeSE"]

	# the dartR package has a bug that doesn't remove NA from standard error
	# filter out non-variant to address this and grab the Fis (doesn't change)
	filt_sub_genl <- gl.filter.monomorphs(filt_sub_genl)
	snp_stats <- gl.report.heterozygosity(filt_sub_genl, n.invariant = 0)
	summary[pop, "Fis"] <- snp_stats[1, "FIS"]
	summary[pop, "Fis_se"] <- snp_stats[1, "FISSE"]
}
```

Write to file
```{r}
write.table(summary, file = paste0("diversity_stats.txt"),
	quote = FALSE, row.names = TRUE)
```

Plot diversity stats to pdf
```{r}
mybarplot <- function(data_val, data_se, names, main) {
	myplot <- barplot(data_val, names = names, las = 2,
		main = main,
		ylim = c(min(c(0, 1.2 * (data_val - data_se * 2))),
			max(0, 1.2 * max(data_val + data_se * 2))))
	arrows(myplot, y0 = data_val + data_se * 2, y1 = data_val - data_se * 2,
		angle = 90, code = 3, length = 0.1)
}

pdf("diversity_stats.pdf", width = 10, height = 10)
mybarplot(summary$Ho_auto, summary$Ho_auto_se, names = rownames(summary),
	main = "Autosomal observed heterozygosity (Ho)")
mybarplot(summary$Ho_SNP, summary$Ho_SNP_se, names = rownames(summary),
	main = "SNP observed heterozygosity (Ho)")
mybarplot(summary$He_auto, summary$He_auto_se, names = rownames(summary),
	main = "Autosomal expected heterozygosity (He)")
mybarplot(summary$He_SNP, summary$He_SNP_se, names = rownames(summary),
	main = "SNP unbiased expected heterozygosity (uHe)")
mybarplot(summary$Fis, summary$Fis_se, names = rownames(summary),
		main = "Inbreeding coefficient Fis\n(1 - Ho / uHe)")
invisible(dev.off())
```
