---
title: "Plot a phylogenetic tree"
output: html_document
---

This document describes the process of plotting a phylogenetic tree  
Initially, I was having trouble reading the complex NEXUS format from IQ-TREE (for concordance factors)  
The function `read.beast()` from package `treeio` associated with package `ggtree` fails  

To get around this, I have written a short python script `concord_to_newick.py`  
Pass the complicated concordance Nexus tree to that script, then use the output for plotting with this script  

The input tree should be in Newick format and can include node labels  
Files are expected to be in the current working directory (set with `setwd("")`)  

The outgroup file has a single tip label per line  
The clade file has a search term per line (or multiple, tab separated) unique for taxa in a clade for collapsing  

Load libraries and read input  
```{r}
suppressMessages(library(ape))
suppressMessages(library(phytools))
treefile <- "phylo.treefile"
concord_treefile <- "concord_scf.tre"
outgroup_file <- "outgroup.txt"
clade_file <- "clades.txt"
```

Read in the clades file if present
```{r}
if (file.exists(clade_file)) {
	clades_present <- TRUE
	clade_lines <- vector(mode = "list")
	index <- 1
	con <- file(clade_file, "r")
	while (TRUE) {
		line <- readLines(con, n = 1)
		if (length(line) == 0) {
			break
		}
		clade_lines[index] <- line
		index <- index + 1
	}
	close(con)
} else {
	clades_present <- FALSE
}
```

Read in the tree file
```{r}
tree <- read.tree(treefile)
```

If there is an outgroup specified, root the tree on the outgroup
```{r}
if (file.exists(outgroup_file)) {
	outgroups <- read.table(outgroup_file)[, 1]
	these_outgroups <- outgroups[outgroups %in% tree$tip.label]
	rootnode <- getMRCA(tree, as.character(these_outgroups))
	position <- 0.5 * tree$edge.length[which(tree$edge[, 2] == rootnode)]
	rooted_tree <- reroot(tree, rootnode, position, edgelabel = TRUE)
	tree <- rooted_tree
}
```

If using site concordance factors, capture them as a dataframe  
(This works if captured with my `concord_to_newick.py` script: "sCF/sDF1/sDF2")
```{r}
if (file.exists(concord_treefile)) {
	concord_tree <- read.tree(concord_treefile)
	# if an outroup is present, root this tree too
	if (file.exists(outgroup_file)) {
		outgroups <- read.table(outgroup_file)[, 1]
		these_outgroups <- outgroups[outgroups %in% concord_tree$tip.label]
		rootnode <- getMRCA(concord_tree, as.character(these_outgroups))
		position <- 0.5 * concord_tree$edge.length[which(concord_tree$edge[, 2] == rootnode)]
		rooted_tree <- reroot(concord_tree, rootnode, position, edgelabel = TRUE)
		concord_tree <- rooted_tree
	}
	concord_df <- data.frame(matrix(ncol = 3, nrow = concord_tree$Nnode))
	index <- 1
	for (label in concord_tree$node.label) {
		if (any(label == "", label == "Root")) {
			values <- c("", "", "")
		} else {
			values <- strsplit(label, split = "\\/")[[1]]
		}
		concord_df[index, 1:3] <- values
		index <- index + 1
	}
	concord_df <- as.data.frame(lapply(concord_df, as.numeric))
}
```


**Tree Plotting**

1) PDF if wanting a simple tree
```{r}
pdf("tree.pdf", width = 16, height = 28, family = "ArialMT")
```

Option 1: plot the tree as a cladogram
```{r}
plot.phylo(ladderize(tree, right = FALSE),
			no.margin = TRUE,
			use.edge.length = FALSE,
			node.depth = 2,
			font = 1,
			edge.width = 3)
```

Option 2: plot the tree with edge lengths
```{r}
plot.phylo(ladderize(tree, right = FALSE),
			no.margin = TRUE,
			font = 1,
			edge.width = 3)
add.scale.bar(x = mean(par("usr")[1:2]), y = par("usr")[3] + 1, font = 1, lwd = 3)
```

If desired, plot node labels (e.g. bootstrap support) on edges
```{r}
drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
```

Another option is to plot node labels on nodes
```{r}
nodelabels(tree$node.label, adj = c(-0.05, 0.5), frame = "none")
```

If desired and present, plot site concordance factors as pie charts on nodes
```{r}
# to catch if there are any all 0
concord_df[which(concord_df[, 1] == 0 & concord_df[, 2] == 0 & concord_df[, 3] == 0), ] <- c(NA, NA, NA)
nodelabels(pie = concord_df,
	piecol = c("white", "grey", "black"),
	cex = 0.5)
```

Stop creating the PDF
```{r}
invisible(dev.off())
```


2) PDF of zoom into the ingroup  
```{r}
pdf("tree_zoom.pdf", width = 11, height = 22, family = "ArialMT")
```

Create the zoom (assuming the labels are clear/unique for the outgroup)
```{r}
zoom(ladderize(tree, right = FALSE),
		grep(paste(outgroups, collapse = "|"), tree$tip.label, invert = TRUE),
		subtree = FALSE,
		col = "black",
		no.margin = TRUE,
		font = 1,
		edge.width = 3)
add.scale.bar(x = mean(par("usr")[1:2]), y = par("usr")[3] + 1, font = 1, lwd = 3)
```

For zooming, it is not clear how to adjust the node labels to display support  
A new plot is made that changes the node numbers to sequential from the order going in  
This method appears to work  
```{r}
# first, determine the edge labels of the tree that is being zoomed into
old_edges <- ladderize(tree, right = FALSE)$edge

# next, find the node number of the MRCA for the ingroup
mrca_node <- getMRCA(tree, grep(paste(outgroups, collapse = "|"), tree$tip.label, invert = TRUE))

# the first time that node shows up in the edges matrix will be the start point for the zoomed tree
index <- grep(mrca_node, old_edges[, 2])
partner_edges <- old_edges[, 2][(index + 1): length(old_edges[, 2])]

# now take those numbers (numbers < Ntip(tree) are tips) and match node labels
# this should be the order of node labels in the zoomed tree
edge_nums <- partner_edges[partner_edges > Ntip(tree)]
new_node_labels <- tree$node.label[edge_nums - Ntip(tree)]

# plot support on edges
drawSupportOnEdges(new_node_labels, adj = c(0.5, -0.5), frame = "none")
```

Stop creating the PDF
```{r}
invisible(dev.off())
```


3) Collapsing clades  
Start an SVG
```{r}
svg("tree.svg", width = 11, height = 22)
```

Perhaps modify N. Cusimano's function (https://www.en.sysbot.bio.lmu.de/people/employees/cusimano/use_r/mapping.R)?  
See also: https://stackoverflow.com/a/34405198  

First, determine the formatting for the plot and plot it (set up coordinates)  
Define a function to display the tree that will have collapsed clades  

Option 1: Cladogram (no branch lengths)
```{r}
collapse_tree <- function(phylo, ...) {
	plot.phylo(ladderize(phylo, right = FALSE),
		no.margin = TRUE,
		use.edge.length = FALSE,
		node.depth = 2,
		font = 1,
		...)
}
```

Option 2: Phylogram (with meaningful branch lengths)
```{r}
collapse_tree <- function(phylo, ...) {
	plot.phylo(ladderize(phylo, right = FALSE),
		no.margin = TRUE,
		font = 1,
		...)
}
```

Option 3: Phylogram, but reducing the impact of long edges (adjust num_edges to fit your tree's long branches)
```{r}
num_edges <- 3
collapse_tree <- function(phylo, ...) {
	plotBreakLongEdges(ladderize(phylo, right = FALSE),
		no.margin = TRUE,
		font = 1,
		n = num_edges,
		...)
}
```

Now run the plotting and collapsing of clades  
```{r}
# check if there are clades present
if (! clades_present) {
	stop(print("There needs to be a clades file specified to collapse clades\n"), call. = FALSE)
}

# plot the tree to get a coordinate system
collapse_tree(tree)

# determine triangle coordinates and edges and tips to drop for clades
tri_coords <- vector(mode = "list", length = length(clade_lines))
tips_drop <- vector(mode = "list", length = length(clade_lines))
edges_drop <- vector(mode = "list", length = length(clade_lines))
labels <- vector(mode = "list", length = length(clade_lines))
root_node <- length(tree$tip.label) + 1

for (index in seq_len(length(clade_lines))) {
	# extract the search terms and determine the taxa and label
	taxa <- vector(mode = "character")
	terms <- strsplit(clade_lines[[index]], split = "\t")[[1]]
	for (term in terms) {
		hits <- tree$tip.label[grep(term, tree$tip.label)]
		taxa <- c(taxa, hits)
	}
	labels[index] <- paste(terms, collapse = "")

	# find most recent common ancestor
	mrca_node <- getMRCA(tree, taxa)	# node number

	# determine the tips that comprise the clade and their plotting numbers
	leaves <- extract.clade(tree, mrca_node)$tip.label
	tips_drop[index] <- list(leaves)
	leaf_nodes <- match(leaves, tree$tip.label)

	# determine the edges that need to be dropped
	edges <- which.edge(tree, leaves)
	edges_drop[index] <- list(edges)

	# determine the clade internal node indices and remove labels
	# but NOT the node at the base of the clade
	mrca_index <- mrca_node - Ntip(tree)
	edge_index <- unique(tree$edge[edges, 1]) - Ntip(tree)
	edge_index <- setdiff(edge_index, mrca_index)
	tree$node.label[edge_index] <- ""

	# determine the coordinate for the mrca node in the plot
	lastplot <- get("last_plot.phylo", envir = .PlotPhyloEnv)
	xcoord1 <- lastplot$xx[mrca_node]
	ycoord1 <- lastplot$yy[mrca_node]

	# determine the coordinates for the base of the triangle
	# x is the max of the clade tips
	xcoord2 <- max(lastplot$xx[leaf_nodes])
	# y coords are the min and max of the clade (traingle may not be even)
	ycoord2 <- min(lastplot$yy[leaf_nodes])
	ycoord3 <- max(lastplot$yy[leaf_nodes])
	# y coords could be the difference between min and max plus/minus from the mrca y
	#spread IS max(lastplot$yy[leaf_nodes]) - min(lastplot$yy[leaf_nodes])
	#ycoord2 IS ycoord1 - (spread / 2)
	#ycoord3 IS ycoord1 + (spread / 2)

	# copy the coords
	xcoords <- c(xcoord1, xcoord2, xcoord2)
	ycoords <- c(ycoord1, ycoord2, ycoord3)
	tri_coords[index] <- list(list(xcoords, ycoords))
}

# now plot the tree with the edges and tips of interest removed
mytree <- tree

# set tip labels to be dropped to NA
all_drops <- unlist(tips_drop)
mytree$tip.label[match(all_drops, mytree$tip.label)] <- NA

# determine the edges in the original tree and their order in the ladderized one
# designate the edges as unique combos of column values in the tree$edge table
all_edges <- unlist(edges_drop)
edge_combos <- tree$edge[all_edges, ]	# the edge order is specific to the unaltered tree
edges_char <- paste0(as.character(edge_combos[, 1]), "_", as.character(edge_combos[, 2]))
new_edges_char <- paste0(as.character(lastplot$edge[, 1]), "_",
	as.character(lastplot$edge[, 2]))

# set the specific edges to have width -1
edge_widths <- rep(3, nrow(mytree$edge))
edge_widths[new_edges_char %in% edges_char] <- -1
collapse_tree(mytree, edge.width = edge_widths)

# add the triangles
for (coordset in tri_coords) {
	polygon(coordset[[1]], coordset[[2]], lwd = 3)
}

# add clade labels
for (index in seq_len(length(labels))) {
	label <- labels[[index]]
	xcoord <- tri_coords[[index]][[1]][3]
	ycoord <- mean(tri_coords[[index]][[2]][2: 3])
	# alternatively, if using the even spread,
	#ycoord IS tri_coords[[index]][[2]][1]
	text(xcoord, ycoord, labels = label, pos = 4)
}
```

If the collapsed clades tree is a phylogram, add a scale bar
```{r}
add.scale.bar(x = mean(par("usr")[1:2]), y = par("usr")[3] + 1, font = 1, lwd = 3)
```

If desired, plot node labels (e.g. bootstrap support) on edges
```{r}
drawSupportOnEdges(tree$node.label, adj = c(0.5, -0.5), frame = "none")
```

Another option is to plot node labels on nodes
```{r}
nodelabels(tree$node.label, adj = c(-0.05, 0.5), frame = "none")
```

If desired and present, plot site concordance factors as pie charts on nodes
```{r}
# to catch if there are any all 0
concord_df[which(concord_df[, 1] == 0 & concord_df[, 2] == 0 & concord_df[, 3] == 0), ] <- c(NA, NA, NA)
nodelabels(pie = concord_df,
	piecol = c("white", "grey", "black"),
	cex = 1)
```

Stop creating the SVG
```{r}
invisible(dev.off())
```
